Experiences with Planning for Natural Language Generation

### 2009-05-26 trunk/Papers/ci-spark/experiments.tex
wrote something about xtag experiment

>>>>

The first experiment already gives us some initial insights into the
appropriateness of planning for the sentence generation domain: On the
examples we looked at, search times were low, but FF and SGPLAN spend
a lot of time on the initial grounding step.  However, one weakness of
this experiment is that it uses a tiny grammar, consisting of just
those 12 lexicon entries that are needed for the experiment. While the
grounding problem can only get worse with larger grammars, the
experiment by itself does not allow us to make clear statements about
the search efficiency.

To address this problem, we ran a second sentence generation
experiment. This time, we used the XTAG Grammar \citep{xtag01:_tr}, a
large-scale TAG grammar of English. XTAG contains lexicon entries for
about 17,000 uninflected words using about 1100 different elementary
trees. Although XTAG does not contain semantic information, it is
possible to automatically equip the lexicon entries with inferred
semantic representations based on the words in the lexicalized
elementary tree. The result is a highly ambiguous grammar: The most
ambiguous word, ``ask'', is the anchor of 314 lexicon entries.

In our experiment, we were especially interested in two
questions. First, how would the planners handle the search problem
involved in generating with such a large and ambiguous grammar?
Second, would it be harder to generate sentences containing verbs with
multiple arguments, given that verbs with more arguments lead to
actions with more parameters and therefore more instances? To answer
these questions, we generated sentences of the form ``S and S and
\ldots and S'', where each S was a sentence; we called the number of
sentences in the conjunction $n$. Each S was a sentence of the form
``the businessman sneezes'', ``the businessman admires the girl'', or
``the businessman gives the girl the book''---that is, they varied in
the number $k$ of syntactic arguments the verb expects (1 for the
intransitive verb ``sneeze'', 2 for the transitive verb ``admire'',
and 3 for the ditransitive verb ``give''). This means that the output
sentence for parameters $n$ and $k$ contained $n(2k+2)-1$ words. The
instances were set up in such a way that the generation of referring
expressions was trivial, so this experiment was purely a surface
realization task.

\begin{figure}
  \centering

  \caption{Results for the XTAG experiment. The horizontal axis
    represents parameters $(k,n)$ from $(1,1)$ to $(3,10)$ in
    lexicographic order. The vertical axis is runtime in milliseconds.}
  \label{fig:xtag-graph}
\end{figure}

Again we ran the experiment with FF 2.3 and SGPLAN 5.2.2; this time,
we also ran the experiment with the Metric-FF implementation of FF
because FF 2.3 had a bug that caused it to switch to a slower search
heuristic for larger problem instances, and which was fixed in
Metric-FF. The results are shown in
Fig.~\ref{fig:xtag-graph}. \todo{blah blah talk about results a bit}
Unlike in the referring expressions experiment above, our own
Graphplan implementation was unusably slow here for serious problem
instances---for $k=1$ and $n=3$, this planner took over two minutes,
and it exceeded its memory limit of 16 GB for $n>3$. \todo{blah blah
  talk about what we learn about FF etc}
